---
title: "Getting Started with ggguides"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ggguides}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(ggguides)
library(ggplot2)
```

## Overview

**ggguides** simplifies common legend operations in ggplot2. Instead of memorizing
theme element names and their nested structure, you can use intuitive one-liner
functions to position, style, and manage legends.

**Key features:**

- Position legends with `legend_left()`, `legend_right()`, `legend_top()`, `legend_bottom()`, `legend_inside()`
- Style legends with `legend_style()` for fonts, backgrounds, and borders
- Wrap legend entries with `legend_wrap()` for multi-column layouts
- Collect legends from patchwork compositions with `collect_legends()`

## Installation

```{r eval=FALSE}
# Install from GitHub
# install.packages("pak")
pak::pak("gcol33/ggguides")
```

## Basic Usage

Let's create a simple plot to demonstrate the legend helpers:
```{r base-plot}
p <- ggplot(mtcars, aes(mpg, wt, color = factor(cyl))) +
  geom_point(size = 3) +
  labs(color = "Cylinders")

p
```

### Moving the Legend

Position the legend on any side with a single function call:

```{r position-left}
p + legend_left()
```

```{r position-bottom}
p + legend_bottom()
```

### Inside Positioning

Place the legend inside the plot area using coordinates or shortcuts:

```{r inside-topright}
p + legend_inside(position = "topright")
```

```{r inside-coords}
p + legend_inside(x = 0.02, y = 0.98, just = c("left", "top"))
```

### Removing the Legend

```{r legend-none}
p + legend_none()
```

## Styling Legends

Use `legend_style()` to customize the legend appearance:

```{r style-size}
p + legend_style(size = 14)
```

```{r style-full}
p + legend_style(
  size = 12,
  title_size = 14,
  title_face = "bold",
  background = "#FFF3E0",
  background_color = "#FF9800"
)
```

## Wrapping Legend Entries

For legends with many entries, use `legend_wrap()` to create multi-column layouts:

```{r wrap-example}
ggplot(mpg, aes(displ, hwy, color = class)) +
  geom_point() +
  legend_wrap(ncol = 2)
```

## Combining Functions

ggguides functions compose naturally with the `+` operator:

```{r combined}
ggplot(mpg, aes(displ, hwy, color = class)) +
  geom_point() +
  legend_left() +
  legend_style(size = 12, title_face = "bold", background = "#FFF3E0")
```

```{r combined-wrap}
ggplot(mpg, aes(displ, hwy, color = class)) +
  geom_point() +
  legend_wrap(ncol = 2) +
  legend_bottom()
```

## What's Next

- See the [Legend Positioning](positioning.html) article for detailed position control
- See the [Styling & Customization](styling.html) article for advanced styling options
- See the [Patchwork Integration](patchwork.html) article for multi-panel workflows
