---
title: "Multiple Legends"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple Legends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(ggguides)
library(ggplot2)

# Theme with transparent backgrounds for pkgdown light/dark mode
theme_set(
  theme_grey() +
    theme(
      plot.background = element_rect(fill = "transparent", color = NA),
      panel.background = element_rect(fill = "transparent", color = NA),
      legend.background = element_rect(fill = "transparent", color = NA),
      legend.key = element_rect(fill = "transparent", color = NA),
      legend.box.background = element_rect(fill = "transparent", color = NA)
    )
)
```

## Overview

When a plot maps multiple aesthetics (colour, size, shape, etc.), ggplot2 creates separate legends for each. ggguides provides functions to control these legends individually:

- **Hide specific legends** with `legend_hide()`

- **Keep only certain legends** with `legend_select()`

- **Control display order** with `legend_order_guides()`

- **Force merge/split** with `legend_merge()` and `legend_split()`

- **Position legends separately** using `by` parameter on position functions

- **Style legends separately** using `by` parameter on `legend_style()`

## Example Plot

```{r example-plot}
# Plot with multiple aesthetics
p <- ggplot(mtcars, aes(mpg, wt,
                        color = factor(cyl),
                        size = hp,
                        shape = factor(am))) +
  geom_point() +
  labs(color = "Cylinders", size = "Horsepower", shape = "Transmission")

p
```

## Hiding Legends

Use `legend_hide()` to remove specific legends while keeping others:

```{r hide-legends}
# Hide the size legend
p + legend_hide(size)

# Hide multiple legends
p + legend_hide(size, shape)
```

## Selecting Legends

Use `legend_select()` to keep only certain legends (inverse of `legend_hide()`):

```{r select-legends}
# Keep only the colour legend
p + legend_select(colour)

# Keep colour and shape
p + legend_select(colour, shape)
```

## Controlling Legend Order

By default, legends appear in an unspecified order. Use `legend_order_guides()` to control the display order:

```{r order-legends}
# Default order
p

# Size legend first, then colour, then shape
p + legend_order_guides(size = 1, colour = 2, shape = 3)
```

## Merging and Splitting Legends

ggplot2 automatically merges legends when they have the same title and matching labels. Use `legend_merge()` and `legend_split()` to override this behavior.

### Forcing Merge

```{r merge-legends}
# Plot where colour and fill map to the same variable
p_merge <- ggplot(mtcars, aes(mpg, wt, color = factor(cyl), fill = factor(cyl))) +
  geom_point(shape = 21, size = 4, stroke = 1.5) +
  labs(color = "Cylinders", fill = "Cylinders")

# Legends merge automatically when titles and labels match
p_merge

# Explicitly request merge (reinforces default behavior)
p_merge + legend_merge(colour, fill)
```

### Forcing Split

```{r split-legends}
# Force separate legends even when they could merge
p_merge + legend_split(colour, fill)
```

## Positioning Legends Separately

Position functions (`legend_left()`, `legend_right()`, `legend_top()`, `legend_bottom()`) accept a `by` parameter to position specific legends:

```{r position-separately}
# Place colour legend on the left, size legend at bottom
p +
  legend_hide(shape) +
  legend_left(by = "colour") +
  legend_bottom(by = "size")
```

```{r position-top-right}
# Colour legend on top, size on right
p +
  legend_hide(shape) +
  legend_top(by = "colour") +
  legend_right(by = "size")
```

## Styling Legends Separately

Use the `by` parameter on `legend_style()` to apply different styles to different legends:

```{r style-separately}
p +
  legend_hide(shape) +
  legend_style(title_face = "bold", background = "grey95", by = "colour") +
  legend_style(size = 10, by = "size")
```

## Four Legends, One per Side

When a plot has four legends and you want one on each side (top, bottom, left, right), you can fine-tune each legend along three axes:

1. **Side placement** — `legend_top/bottom/left/right(by = "<aes>")`

2. **Distance from the panel** — `legend_style(by = "<aes>", margin = c(t, r, b, l))`

3. **Slide along the side** — `legend_style(by = "<aes>", justification = ...)`

For top/bottom legends, `justification` is `"left"`, `"center"`, `"right"` (or a number in `[0, 1]`). For left/right legends, it's `"top"`, `"center"`, `"bottom"` (or a number).

```{r four-sides, fig.width = 8, fig.height = 6}
p4 <- ggplot(mtcars, aes(mpg, wt,
                         colour = factor(cyl),
                         fill   = factor(gear),
                         size   = hp,
                         shape  = factor(am))) +
  geom_point(stroke = 1.2) +
  labs(colour = "Cyl", fill = "Gear", size = "HP", shape = "AM")

p4 +
  # 1. Send each legend to its side
  legend_top   (by = "colour") +
  legend_bottom(by = "fill")   +
  legend_left  (by = "size")   +
  legend_right (by = "shape")  +

  # 2. Slide each legend along its side
  legend_style(by = "colour", justification = "left") +
  legend_style(by = "fill",   justification = "right") +
  legend_style(by = "size",   justification = "top") +
  legend_style(by = "shape",  justification = "bottom") +

  # 3. Nudge each legend toward/away from the panel via margin (cm)
  legend_style(by = "colour", margin = c(0, 0, 0.3, 0)) +
  legend_style(by = "fill",   margin = c(0.3, 0, 0, 0)) +
  legend_style(by = "size",   margin = c(0, 0.3, 0, 0)) +
  legend_style(by = "shape",  margin = c(0, 0, 0, 0.3))
```

Each `legend_style(by = ...)` call is additive — you can chain as many as you need to tune one legend at a time without affecting the others.

## Combining Multiple Controls

All functions work together:

```{r combined}
# Complex example: hide shape, position colour on left with bold title,
# position size at bottom with smaller text
p +
  legend_hide(shape) +
  legend_left(by = "colour") +
  legend_style(title_face = "bold", title_size = 14, by = "colour") +
  legend_bottom(by = "size") +
  legend_style(size = 9, direction = "horizontal", by = "size")
```

## Summary

| Function | Purpose | Parameters |
|----------|---------|------------|
| `legend_hide()` | Hide specific legends | Aesthetic names (unquoted) |
| `legend_select()` | Keep only specific legends | Aesthetic names (unquoted) |
| `legend_order_guides()` | Control legend display order | Named args: `aes = order` |
| `legend_merge()` | Force legends to merge | Aesthetic names (unquoted) |
| `legend_split()` | Force legends to stay separate | Aesthetic names (unquoted) |
| `legend_left(by=)` | Position one legend on left | `by = "aesthetic"` |
| `legend_style(by=)` | Style one legend | `by = "aesthetic"` + style args |

**Learn more:**

- [Legend Positioning](positioning.html) for single-legend placement

- [Styling & Customization](styling.html) for legend appearance

- [Patchwork Integration](patchwork.html) for multi-panel plots
