---
title: "Patchwork Integration"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Patchwork Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(ggguides)
library(ggplot2)
```

## Overview

When combining multiple plots with patchwork, legends often need special handling:

- **Duplicate legends** - Same aesthetic mapped in multiple plots creates redundancy
- **Alignment issues** - Legends may not align properly across panels
- **Spanning** - Legends should sometimes span multiple rows

ggguides provides `collect_legends()` and `align_guides_h()` to address these challenges.

## The Problem: Duplicate Legends

```{r problem, fig.height=4}
library(patchwork)

p1 <- ggplot(mtcars, aes(mpg, wt, color = factor(cyl))) +
  geom_point() + labs(title = "Plot 1", color = "Cylinders")
p2 <- ggplot(mtcars, aes(mpg, hp, color = factor(cyl))) +
  geom_point() + labs(title = "Plot 2", color = "Cylinders")

# Default patchwork: duplicate legends
p1 | p2
```

## Basic Legend Collection

Use `collect_legends()` to gather legends from all plots in a composition:

```{r collect-basic, fig.height=4}
collect_legends(p1 | p2)
```

### Position Options

Control where the collected legend appears:

```{r collect-bottom, fig.height=5}
collect_legends(p1 | p2, position = "bottom")
```

```{r collect-left, fig.height=4}
collect_legends(p1 | p2, position = "left")
```

## Stacked Plots

For vertically stacked plots, legends can be centered or span the full height.
Using different plot heights makes the spanning behavior more visible.

### Default: Centered

```{r stacked-default, fig.height=7}
# Create plots with different heights using plot_layout
p3 <- ggplot(mtcars, aes(mpg, disp, color = factor(cyl))) +
  geom_point() + labs(title = "Plot 3", color = "Cylinders")

# Stack with different heights: 1, 1/2, 1/4
stacked <- (p1 / p2 / p3) + plot_layout(heights = c(4, 2, 1))
collect_legends(stacked, position = "right")
```

### Spanning Full Height

Use `span = TRUE` to make the legend fill the full height:

```{r stacked-span, fig.height=7}
gt <- collect_legends(stacked, position = "right", span = TRUE)
grid::grid.draw(gt)
```

### Attaching to Specific Rows

Attach the legend to specific row(s) instead of spanning all:

```{r span-row1, fig.height=7}
# Legend attached to row 1 only (the tallest plot)
gt <- collect_legends(stacked, position = "right", span = 1)
grid::grid.draw(gt)
```
```{r span-row12, fig.height=7}
# Legend attached to rows 1 and 2
gt <- collect_legends(stacked, position = "right", span = 1:2)
grid::grid.draw(gt)
```

## Horizontal Guide Alignment

When plots have different widths due to axis labels, use `align_guides_h()` to
align them horizontally:

```{r align-guides, fig.height=5}
# Plots with different y-axis label widths
p_short <- ggplot(mtcars, aes(mpg, wt)) +
  geom_point() + labs(y = "wt")

p_long <- ggplot(mtcars, aes(mpg, disp)) +
  geom_point() + labs(y = "Displacement (cu.in.)")

# Without alignment
p_short / p_long
```

```{r align-guides-fixed, fig.height=5}
# With horizontal alignment
align_guides_h(p_short / p_long)
```

## Combining with Styling

ggguides functions work together:

```{r combined, fig.height=5}
p1_styled <- p1 + legend_style(size = 11, title_face = "bold")
p2_styled <- p2 + legend_style(size = 11, title_face = "bold")

collect_legends(p1_styled | p2_styled, position = "right")
```

## Complex Layouts

Handle more complex patchwork layouts:

```{r complex, fig.height=6}
p4 <- ggplot(mtcars, aes(qsec, wt, color = factor(cyl))) +
  geom_point() + labs(title = "Plot 4", color = "Cylinders")

# 2x2 grid
layout <- (p1 | p2) / (p3 | p4)
collect_legends(layout, position = "right")
```

## cowplot Users

For cowplot workflows, consider the [lemon](https://github.com/stefanedwards/lemon) package which provides:

- `g_legend()` - Extract legend as grob
- `grid_arrange_shared_legend()` - Combine plots with shared legend
- `reposition_legend()` - Place legend inside panels

ggguides functions like `legend_style()`, `legend_wrap()`, and position helpers
work on individual plots regardless of layout package.

## Summary

| Function | Purpose | Key Parameters |
|----------|---------|----------------|
| `collect_legends()` | Gather legends from patchwork | `position`, `span` |
| `align_guides_h()` | Align plots horizontally | None |

**Learn more:**

- [Legend Positioning](positioning.html) for single-plot placement
- [Styling & Customization](styling.html) for legend appearance
